---
format:
  revealjs:
    footer: "Packages for forecasting --- [cmu-delphi.github.io/midas-cdc-2023-demo]{.smaller}"
    logo: "gfx/delphi.jpg"
    embed-resources: false
    chalkboard: true
    width: 1280
    height: 720
    theme: [default, themer.scss]
    fig-format: svg
    html-math-method: mathjax
execute:
  cache: true
editor: source
---

```{r setup, include=FALSE}
pak::pkg_install("epidatr")
library(epidatr)
library(dplyr)
```

## What is `epidatr`?

The [Delphi Epidatr package](https://cmu-delphi.github.io/epidatr/) is an R front-end for the [Delphi Epidata API](https://cmu-delphi.github.io/delphi-epidata/), which provides real-time access to epidemiological surveillance data for influenza, COVID-19, and other diseases for the USA at various geographical resolutions, both from official government sources such as the [Center for Disease Control (CDC)](https://www.cdc.gov/datastatistics/index.html) and [Google Trends](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/google-symptoms.html) and private partners such as [Facebook](https://delphi.cmu.edu/blog/2020/08/26/covid-19-symptom-surveys-through-facebook/) and [Change Healthcare](https://www.changehealthcare.com/). It is built and maintained by the Carnegie Mellon University [Delphi research group](https://delphi.cmu.edu/).

This package is designed to streamline the downloading and usage of data from the [Delphi Epidata
API](https://cmu-delphi.github.io/delphi-epidata/). It provides a simple R interface to the API, including functions for downloading data, parsing the results, and converting the data into a tidy format. The API stores a historical record of all data, including corrections and updates, which is particularly useful for accurately backtesting forecasting models. We also provide packages for downstream data processing ([epiprocess](https://github.com/cmu-delphi/epiprocess)) and modeling ([epipredict](https://github.com/cmu-delphi/epipredict)).


## Installation

You can install the stable version of this package from CRAN:

```{r, eval=FALSE}
install.packages("epidatr")
pak::pkg_install("epidatr")
renv::install("epidatr")
```

Or if you want the development version, install from GitHub:

```{r, eval=FALSE}
# Install the dev version using `pak` or `remotes`
pak::pkg_install("cmu-delphi/epidatr")
remotes::install_github("cmu-delphi/epidatr")
renv::install("cmu-delphi/epidatr")
```

## API key setup

The Delphi API requires a (free) API key for full functionality. To generate
your key, register for a pseudo-anonymous account
[here](https://api.delphi.cmu.edu/epidata/admin/registration_form) and see more
discussion on the [general API
website](https://cmu-delphi.github.io/delphi-epidata/api/api_keys.html). See the
`set_api_key()` function documentation for details on how to use your API key.

Note that the private endpoints (i.e. those prefixed with `pvt_`) require a
separate key that needs to be passed as an argument. These endpoints require
specific data use agreements to access.

## Accessing HHS data

```{r}
# Obtain 7-day average of confirmed COVID-19 hospital admissions from the
# Department of Health & Human Services for the US
epidata <- pub_covidcast(
  source = "hhs",
  signals = "confirmed_admissions_covid_1d_7dav",
  geo_type = "nation",
  time_type = "day",
  geo_values = "us",
  time_values = epirange(20230101, 20230601)
)
epidata
```

## Commonly-accessed data sources

## Other useful data sources

Exclusive data including `chng` and `doctor-visits`

## Other endpoints

More than just COVID data!

```{r, eval=FALSE}
avail_endpoints()
```

```{r, echo=FALSE}
avail_endpoints() %>% filter(startsWith(Endpoint, "pub_"), !startsWith(Endpoint, "pub_covid"))
```


