---
format:
  revealjs:
    footer: "Packages for forecasting --- [cmu-delphi.github.io/midas-cdc-2023-demo]{.smaller}"
    logo: "gfx/delphi.jpg"
    embed-resources: false
    chalkboard: true
    width: 1280
    height: 720
    theme: [default, themer.scss]
    fig-format: svg
    html-math-method: mathjax
execute:
  cache: true
editor: source
include-in-header:
  - text: |
      <style type="text/css">
      ul li ul li {
        font-size: 0.85em;
      }
      </style>
---


##  Finding, fetching, and processing epidemiological data with `{epidatr}` and `{epiprocess}`

<br>

#### Delphi Research Group at CMU
#### Slides: Nat DeFries, Dmitry Shemetov, Logan Brooks, others on Delphi tooling team

CDC and MIDAS Forecasting Meeting --- 21 November 2023

Slides are online at <https://cmu-delphi.github.io/midas-cdc-2023-demo>

```{r}
#| include: false

# setup
pak::pkg_install("epidatr")
library(epidatr)
library(dplyr)

compact_print_tbl <- function(df) {
  # Drop extraneous tibble messages, from
  # https://stackoverflow.com/a/64747799/14401472
  #   - first position stores "tibble <size>" statement
  #   - third position stores column classes
  cat(format(df)[c(-1L, -3L)], sep = "\n")
}
```


## The [Delphi `{epidatr}` package](https://cmu-delphi.github.io/epidatr/) is a new R front-end for the [Delphi Epidata API](https://cmu-delphi.github.io/delphi-epidata/)

- streamlines downloading and usage of data from the [Delphi Epidata
API](https://cmu-delphi.github.io/delphi-epidata/)
  - real-time access to epidemiological surveillance data for influenza, COVID-19, and other diseases
  - data from both official government sources such as the [CDC](https://www.cdc.gov/datastatistics/index.html) and from private partners
  - a historical record of all data available, including corrections and updates, which is useful for backtesting of forecasting models.
- provides a simple R interface to the API, with functions for downloading data, parsing results, and converting to tidy format.
  - the `{epidatr}` package is a complete rewrite of the [`{covidcast}` package](https://cmu-delphi.github.io/covidcast/covidcastR/) and `delphi_epidata.R` script, with a focus on speed, reliability, and ease of use
  - the {`covidcast`} package and `delphi_epidata.R` script are deprecated and will no longer be updated


## Conveniently install in the normal ways

- You can install the stable version of this package from CRAN:

```{r}
#| eval: false
#| echo: true
install.packages("epidatr")
pak::pkg_install("epidatr")
renv::install("epidatr")
```

- Or if you want the development version, install from GitHub:

```{r}
#| eval: false
#| echo: true
pak::pkg_install("cmu-delphi/epidatr@dev")
remotes::install_github("cmu-delphi/epidatr", ref = "dev")
renv::install("cmu-delphi/epidatr@dev")
```

- `{epidatr}` requires a (free) API key for full functionality
  - To generate your key, [register](https://api.delphi.cmu.edu/epidata/admin/registration_form) for a pseudo-anonymous account (see the [general API website](https://cmu-delphi.github.io/delphi-epidata/api/api_keys.html) for details) and use `save_api_key()` for help storing the key.
  - (Note: we also have private endpoints (those prefixed with `pvt_`) that require a separate key to be passed as an argument. These endpoints require data use agreements to access.)

:::


## Example: HHS/NHSN hospitalization data

- Fetch national COVID-19 hospital admissions:

```{r}
#| echo: true
#| eval: true
epidata <- pub_covidcast(
  source = "hhs",
  signals = "confirmed_admissions_covid_1d",
  geo_type = "nation",
  time_type = "day",
  geo_values = "us",
  time_values = epirange("2023-01-01", "2023-06-01")
  # (by default, fetches the current version)
)
# `epidata` looks like:
```

```{r cache.extra = epidata}
#| echo: false
#| eval: true
epidata %>%
  select(geo_value, signal, source, geo_type, time_type, time_value, issue, lag, value, everything()) %>%
  print(n = 5L)
```

<!-- ## Example: HHS/NHSN hospitalization data -->

<!-- ```{r} -->
<!-- #| echo: false -->
<!-- #| eval: true -->
<!-- epidata %>% -->
<!--   select(geo_value, signal, source, geo_type, time_type, time_value, issue, lag, value, everything()) %>% -->
<!--   print(n = 5L) -->
<!-- ``` -->

<!-- - Note the key columns: `source, signal, geo_value, time_value` -->
<!-- - `issue` is the date when the row started being valid (when it entered our database^[Usually; exceptions include adding back issues for a new data set and patching to better match upstream history.]) -->


## Example: versioned HHS/NHSN hospitalization data

- Fetch what the query should have looked like back in June ("as of" June 1st):

```{r}
#| echo: true
#| eval: true
epidata <- pub_covidcast(
  source = "hhs",
  signals = "confirmed_admissions_covid_1d",
  geo_type = "nation",
  time_type = "day",
  geo_values = "us",
  time_values = epirange("2023-01-01", "2023-06-01"),
  as_of = "2023-06-01"
)
# `epidata` looks like:
```

```{r cache.extra = epidata}
#| echo: false
#| eval: true
epidata %>%
  select(geo_value, signal, source, geo_type, time_type, time_value, issue, lag, value, everything()) %>%
  print(n = 5L)
```

<!-- - Note that we have fewer rows now, reflecting the reality of data latency, and that the counts are slightly different, reflecting data revisions. -->


## Access other useful data, including Delphi-exclusive sources

- [`chng`](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/chng.html) from [Change Healthcare](https://www.changehealthcare.com/)
    - Outpatient COVID and influenza signals from insurance claims
- [`doctor-visits`](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/doctor-visits.html) from health system partners
    - COVID signals from insurance claims
- [`fb-survey`](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/fb-survey.html) from (parts of) the [COVID-19 Trends and Impact Survey](https://delphi.cmu.edu/blog/2020/08/26/covid-19-symptom-surveys-through-facebook/) in collaboration with Facebook
    - Signals of public behaviors and beliefs about public health during the pandemic

See also `covidcast_epidata()` or the [COVIDcast web site](https://cmu-delphi.github.io/delphi-epidata/api/covidcast_signals.html) for a listing of other COVIDcast data available.

## Access more than just COVID data!

Using `avail_endpoints()` you can find a listing of our other endpoints that serve a wide variety of public health data.  Here we've filtered to non-COVID-specific data.

```{r}
#| eval: true
#| echo: false
invisible(capture.output(endpts <- avail_endpoints()))
endpts %>%
  filter(startsWith(Endpoint, "pub_"), !startsWith(Endpoint, "pub_covid")) %>%
  compact_print_tbl()
```


## Consider subscribing to the [Delphi API mailing list](https://lists.andrew.cmu.edu/mailman/listinfo/delphi-covidcast-api) to be notified of package updates, new data sources, corrections, and other updates


## The `{epiprocess}` package helps work with epidemic datasets

```{r}
#| include: false

# setup
pak::pkg_install("cmu-delphi/epiprocess@main")
library(epiprocess)
```

```{r}
#| eval: false
#| echo: true
pak::pkg_install("cmu-delphi/epiprocess@main")
```

- provides common data structures for epidemiological data sets measured over space and time
- provides utilities for basic signal processing tasks


## `epi_df`: a snapshot of epidata in time

- represents the most up-to-date values of dataset as of a given time
- a subclassed tibble with two required columns: `geo_value` and `time_value`
- and associated metadata: `geo_type`, `time_type`, `other_keys`, `as_of`
- can have any number of other columns, which we call signal (or measured) variables

## `epi_df`: a snapshot of epidata in time

Produce an `epi_df` from `epidatr` output like so:

```{r}
#| echo: true
#| eval: true
tbl <- pub_covidcast(
  source = "hhs",
  signals = "confirmed_admissions_covid_1d",
  geo_type = "state",
  time_type = "day",
  geo_values = "ca,fl,ny,tx",
  time_values = "*"
)
epi_df <- tbl %>%
  # Add NAs to fill gaps, cover same time range for each geo:
  tidyr::complete(geo_value, time_value = tidyr::full_seq(time_value, period = 1L)) %>%
  as_epi_df(
    geo_type = "state",
    time_type = "day",
    as_of = max(tbl$issue)
  )
```

## `epi_df`: a snapshot of epidata in time

```{r cache.extra = epi_df}
#| echo: true
epi_df
```

## `epi_archive`: a collection of historical epidata

- represents the most up-to-date values of dataset as of various given times
- required input columns: `geo_value`, `time_value`, `version`
- can have any number of other key or signal/measured columns

## `epi_archive`: a collection of historical epidata

```{r}
#| echo: true
#| eval: true
tbl <- pub_covidcast(
  source = "hhs",
  signals = "confirmed_admissions_covid_1d",
  geo_type = "state",
  time_type = "day",
  geo_values = "ca,fl,ny,tx",
  time_values = "*", # "*" = all time values
  issues = epirange("1234-01-01", "2023-06-01") # start of range must be before data set start
)
epi_archive <- tbl %>%
  select(
    geo_value, time_value,
    version = issue, admissions = value
  ) %>%
  # don't try to `complete` here; `complete` after `epix_as_of` or inside `epix_slide` computations
  as_epi_archive(compactify = TRUE)
```

## `epi_archive`: a collection of historical epidata

 <!-- cache.extra = epi_archive generates error -->
```{r}
#| echo: true
epi_archive
```

## `epi_archive`: a collection of historical epidata

 <!-- cache.extra = epi_archive generates error -->
```{r}
#| echo: true
epi_archive$DT
```

## Some `epi_slide` use cases

- Calculate rolling or running averages, sums, other statistics
- Calculate custom growth rates, categorical trend definitions, smoothing (see
  also `epiprocess::growth_rate`, `epipredict::step_lag_difference`)
- (Perform latency&revision-naive forecaster backtesting)

## Some `epix_as_of`, `epix_slide` use cases

- Better backtesting: generate pseudoprospective forecasts
- Plot past forecast against data available at generation time
- Plot evolution of how a time series was reported
- Analyze reporting latency, revision behavior, trends
- Improve forecasts when revisions are significant: prepare "version-analogous" training set predictor data

## `epi_df` and `epi_archive` utilities

- `epi_df`
  - `group_by()` - standard grouped operations
  - `epi_slide()` - perform (grouped) time-window computations on an `epi_df`
  - `epi_cor()` - compute correlations between variables in an `epi_df`
- `epi_archive`
  - `epix_merge()` - merge/join two `epi_archive` objects
  - `epix_as_of()` - generate a snapshot `epi_df` from an `epi_archive` object
  - `group_by()` - standard grouped operations
  - `epix_slide()` - perform (grouped) time-windowed computations on several versions
- And more, including outlier detection^[functionality contributed by Evan Ray], growth rate calculation.

## Resources

- [epidatr (stable) documentation](https://cmu-delphi.github.io/epidatr/index.html)
- [epidatr (dev) documentation](https://cmu-delphi.github.io/epidatr/dev/index.html)
- [COVIDcast data source listing](https://cmu-delphi.github.io/delphi-epidata/api/covidcast_signals.html)
- [Other Delphi Epidata API endpoints](https://cmu-delphi.github.io/delphi-epidata/api/README.html)
- [epiprocess documentation](https://cmu-delphi.github.io/epiprocess/index.html)
- [Delphi Epidemiological Forecasting Book](https://cmu-delphi.github.io/delphi-tooling-book/epiprocess.html)
